---
title: "Initial Survey for Mangrove Kingfisher"
author: "Raphaël Nussbaumer <raphael.nussbaumer@arocha.org>"
output: 
  rmdformats::material
  html_document:
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(purrr)
library(lubridate)
library(leaflet)
library(sp)
library(RColorBrewer)
library(DT)
library(leaflet.extras)

project.root = 'C:/Users/rnussba1/ARK/Science - Documents/04 Geolocator/7-Output/GBIF'
export = "2021-05-28"
```

## Load survey


```{r}
surveys <- read_csv(paste0(project.root,"/data/complied_files/export_", export, "_surveys.csv"),
                    col_types = cols(
                      filename = col_character(),
                      observer = col_character(),
                      date = col_date(format = ""),
                      sessionID = col_character(),
                      start_time = col_time(format = ""),
                      end_time = col_time(format = ""),
                      gap_min = col_double(),
                      start_hour = col_time(format = ""),
                      surveyID = col_character()
                    )
                    ) %>% 
  mutate(
    duration = difftime(end_time,start_time, units="mins")-gap_min
  )
records <- read_csv(paste0(project.root,"/data/complied_files/export_", export, "_records.csv"),
                    col_types = cols(
                    specieName = col_character(),
                    specieID = col_double(),
                    date = col_datetime(format = ""),
                    lat = col_double(),
                    lng = col_double(),
                    note = col_character(),
                    recordID = col_character(),
                    surveyID = col_character(),
                    sessionID = col_character()
                  )) 
```

The current dataset includes **`r length(surveys)` surveys**, spanning from `r first(surveys$date)` to `r last(surveys$date)`, totalizing **`r as.integer(sum(surveys$duration/60))` hours**.

## Read data for each survey

Based on the metadata of each survey loaded in the previous section, we can now create the database including all sightings (i.e. individual observation) and associated survey information. 
```{r}
datatable(records, filter = "top", extensions = 'Buttons', options = list(dom = 'Bfrtip', buttons = c('csv')), rownames = FALSE, class = "compact")
```


# Map surveys track and MK locations

```{r}

m <- leaflet(width = "100%") %>%
  addProviderTiles("MapBox", options = providerTileOptions(
    id = "mapbox.satellite",
    accessToken = 'pk.eyJ1IjoicmFmbnVzcyIsImEiOiIzMVE1dnc0In0.3FNMKIlQ_afYktqki-6m0g')) %>% 
  addFullscreenControl()

factpal <- colorFactor(brewer.pal(12, 'Paired'), as.factor(records$surveyID))

for (s in unique(records$surveyID)){
  m <- addPolylines(m, data=records %>% 
                      filter(surveyID==s) %>% 
                      mutate(      
                        day = format(date,"%Y%m%d"),
                        label = paste0(
                         "<b>Date:</b> ", date,'<br>',
                         "<b>surveyID</b>: ", surveyID,'<br>'
                      )),
                    lng = ~lng, lat = ~lat , group = ~sessionID, color=~factpal(surveyID), popup = ~label)
}

m <- addCircleMarkers(m, data = records %>%
                  mutate(
                    day = format(date,"%Y%m%d"),
                    label = paste0(
                       "<b>recordID</b>: ", recordID,'<br>',
                       "<b>Species:</b> ", specieName,'<br>',
                       "<b>Date:</b> ", date, '<br>',
                       "<b>surveyID</b>: ", surveyID,'<br>'
                  )),
                lng = ~lng, lat = ~lat, popup = ~label,
                radius = ~ifelse(specieID==400, 10, 6),
                stroke = FALSE,
                fillOpacity = ~ifelse(specieID==400, 1, 0.8),
                color = ~factpal(surveyID),
                group = ~sessionID)

m <- m %>% addLayersControl(
    overlayGroups =  records$sessionID, #format(d$date,"%Y%m%d"),
    options = layersControlOptions(collapsed = FALSE)
  )

m
```

## All MK sightings

```{r}
dMK <- read_csv(paste(project.root,'/4-Data/BL_EVENT_mangrovekingfisherproj.csv',sep=""), 
         col_types = cols_only(
      date = col_date(format = ""),
      time = col_time(format = ""),
      locationLat = col_double(),
      locationLon = col_double(),
      lead = col_character()
    )) %>% 
  mutate(lead = gsub("\\@.*","",lead))

datatable(dMK, filter = "top", extensions = 'Buttons', options = list(dom = 'Bfrtip', buttons = c('csv')), rownames = FALSE, class = "compact")
```

           
```{r}

m <- leaflet(width = "100%") %>%
  addProviderTiles("MapBox", options = providerTileOptions(
    id = "mapbox.satellite",
    accessToken = 'pk.eyJ1IjoicmFmbnVzcyIsImEiOiIzMVE1dnc0In0.3FNMKIlQ_afYktqki-6m0g')) %>% 
  addFullscreenControl()

factpal <- colorFactor(brewer.pal(12, 'Paired'), as.factor(d$survey_ID))

for (s in unique(d$survey_ID)){
  m <- addPolylines(m, data=d %>% 
                      filter(survey_ID==s),
                    lng = ~Longitude, lat = ~Latitude , group = ~Date, color='black')
}

m <- addMarkers(m, data = records %>% 
                        filter(specieID==400) %>% 
                  mutate(label = paste(sep = "",
                                   "<b>Date:</b> ", date,'<br>',
                                   "<b>Note</b>: ", note
                  )), 
                lng = ~lng, lat = ~lat, popup = ~label,
                group = 'from surveys'
                )

m <- addMarkers(m, data = dMK %>% 
                  mutate(label = paste(sep = "",
                                   "<b>Date:</b> ", date, ' ', time,'<br>',
                                   "<b>Observer:</b> ", lead,'<br>'
                  )), 
                lng = ~locationLon, lat = ~locationLat, popup = ~label,
                group = 'from KBM',
                clusterOptions = markerClusterOptions()
                )

m <- m %>% addLayersControl(
    overlayGroups = c('from KBM','from surveys'),
    options = layersControlOptions(collapsed = FALSE)
  )

m
```
## Bird diversity

```{r}
records %>% 
  group_by(specieName) %>% 
  filter(specieName!= 'Unidentified') %>% 
  summarise(
    n = n(),
    occurance = n/length(unique(records$survey_ID)),
    #avgtime = parse_time(format(mean(as.POSIXct(Time, format="%H:%M:%S"))), format="%Y-%M-%D %H:%M:%S"),
    .groups = "drop"
    ) %>% 
  arrange(desc(n)) %>% 
  datatable( filter = "top", extensions = 'Buttons', options = list(dom = 'Bfrtip', buttons = c('csv')), rownames = FALSE, class = "compact")
```











